---
title: 'MSDS 597 Assignment #7'
author: "Zach Wilson"
date: "November 7, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

As usual, the first step in this process is to load all the necessary libraries.

```{r, echo = TRUE, include = TRUE} 
# load necessary libraries
library(tidyverse)
library(stringr)
library(broom)
library(viridis)
library(choroplethr)
```

Once that is complete, the maternal mortality data is read in and cleaned up.  This clean up included removing the spaces used to indicate where commas would have been in the numbers, creating new lower and uper maternal mortality bound colums from the information in square brackets in the original maternal mortality data, converting the relevant colums from characters to numbers, and creating new rows for the skilled health percentage rows that gave year ranges.

```{r, echo = TRUE, include = TRUE} 
# read in and tidy up the maternal mortality data
print("debug 1")
mm <- read_csv("../data/xmart.csv")
names(mm) <- c("country", "year", "maternal_mortality", "skilled_health_percentage")


# remove spaces in maternal mortality numbers where commas should be
mm$maternal_mortality <- str_replace_all(mm$maternal_mortality, "[\\s]+", "")

# turn the numbers inside square brackets into lower bounds and upper bounds
  # first grab the whole mm string when its there and remove the [, -, and ]
tmp <- str_split(mm$maternal_mortality, "[\\[\\-\\]]")

  # get rid of all but the first number in the maternal mortality column
mm$maternal_mortality <- str_replace(mm$maternal_mortality, "[\\[][0-9]*[-][0-9]*[\\]]", "")

  # create the upper and lower bound lists
mm_lower_bound <- vector(length = length(tmp), mode = "character")
mm_upper_bound <- vector(length = length(tmp), mode = "character")
i = 1
for (ss in tmp) {
  mm_lower_bound[i] = ss[2]
  mm_upper_bound[i] = ss[3]
  i <- i + 1
}
 
  # add the maternal mortality lower and upper bounds columns to the tibble
mm$mm_lower_bound <- mm_lower_bound
mm$mm_upper_bound <- mm_upper_bound
# mm <- mutate(mm, mm_lower_bound)
# mm <- mutate(mm, mm_upper_bound)

  # convert the maternal mortality columns to numeric
mm$maternal_mortality <- as.numeric(mm$maternal_mortality)
mm$mm_lower_bound <- as.numeric(mm$mm_lower_bound)
mm$mm_upper_bound <- as.numeric(mm$mm_upper_bound)

# create a seperate row for each year of skilled health care percentage data
year_range_loc = str_detect(mm$year, "-")
mm_new <- data.frame(country = character(), year = character(), maternal_mortality = double(),
                     mm_lower_bound = double(), mm_upper_bound = double())
i = 1
for (y in year_range_loc){
  if (year_range_loc[i]) {
    ys <- str_split(mm$year[i], "-")
    mm$year[i] <- ys[[1]][1]
    newrow <- mm[i,]
    newrow$year <- ys[[1]][2]
    mm_new <- rbind(mm_new, newrow)
  }
  i = i + 1
}
mm <- rbind(mm, mm_new) 
  
# make the year column numeric
mm$year <- as.integer(mm$year)
```

Once the maternal mortality data has been read in and cleaned up, the gross national income is read in.  This data is not in a tidy format and so once the data has been cleaned up a bit a gather is run to create a data frame with only three columns, one of which is year.

```{r, echo = TRUE, include = TRUE} 
# read in and tidy up the gross national income data
gni <- read_csv("../data/WHS9_93.csv", skip = 1)

# get rid of the space where the comma should be in the 2013 column and convert column to integer
gni$`2013` <- str_replace_all(gni$`2013`, "[\\s]+", "")
gni$`2013` <- as.integer(gni$`2013`)


# reformat so that there is one entry per country per year
gni <- gather(gni, `2013`:`2012`:`2011`:`2010`:`2009`:`2008`:`2007`:`2006`:`2005`:`2004`:`2003`:
              `2002`:`2001`:`2000`:`1999`:`1998`:`1997`:`1996`:`1995`:`1994`:`1993`:`1992`:
              `1991`:`1990`, key = "year", value = gross_nat_income)

# make year a numeric column
gni$year <- as.integer(gni$year)

# get rid of any rows that gave NAs for gross national income
gni <- na.omit(gni)
```

Finally, these two data sets are joined using country and year as the primary keys.

```{r, echo = TRUE, include = TRUE} 
# perform a join on the gni and mm tibbles using year and country as the keys
mm_gni_clean <- inner_join(mm, gni, by = c("country" = "Country", "year" = "year"))
```

Now a plot showing the relationship between income and and maternal mortality is created

```{r, echo = TRUE, include = TRUE} 
# plot maternal mortality vs income
mm_gni_clean %>%
  ggplot(aes(x = gross_nat_income, y = maternal_mortality)) + 
  geom_point(show.legend = TRUE) + 
  ggtitle("Maternal Mortality vs Gross National Income") +
  labs(x = "Per Capita Gross National Income ($)", 
       y = "Maternal Mortality Ratio (per 100,000 live births)") +
  scale_color_viridis(discrete = TRUE) +
  theme_bw()
```

```{r, echo = TRUE, include = TRUE} 
# plot maternal mortality vs income
mm_gni_clean %>%
  ggplot(aes(x = gross_nat_income, y = skilled_health_percentage)) + 
  geom_point(show.legend = TRUE) + 
  ggtitle("Percentage of Births Attended by Skilled Health Care vs Gross National Income") +
  labs(x = "Per Capita Gross National Income ($)", 
       y = "Percentage of Births Attended by Skilled Health Care") +
  scale_color_viridis(discrete = TRUE) +
  theme_bw()
```